---
title: "Reproducible Research: Peer Assessment 1 (Elisa Du)"
output: 
  html_document:
    keep_md: true
---
Load some packages.
```{r Packages, echo=TRUE,results='hide'}
library(knitr)
library(dplyr)
library(ggplot2)
```
  
First, set global options.
```{r setglobal}
opts_chunk$set(echo=TRUE,cache=FALSE)
```

## Loading and preprocessing the data
```{r LoadData}
df<-read.csv('activity.csv',sep=',',na.strings = 'NA')
head(df)
```

## What is mean total number of steps taken per day?  
To make a histogram of total number of steps taken each day, first sort data by date.
```{r sortByDate}
df2<-df[!is.na(df$steps),] # subset all rows with non-NA values
head(df2)
df_GroupByDate<-group_by(df2,date) # group by 'date' as factor
head(df_GroupByDate)
df_stepsPerDay<-summarise(df_GroupByDate,TotalSteps=sum(steps)) # create new data frame
head(df_stepsPerDay)
```
Now we can plot the histogram.     
```{r histogram,fig.height= 4}
plot1<-qplot(TotalSteps,data=df_stepsPerDay,geom='histogram')
plot1+labs(x='Total Steps Taken per Day')
```  
  
Now we find the mean total number of steps taken for each day.    
```{r mean}
meanTotalSteps<- mean(df_stepsPerDay$TotalSteps)
meanTotalSteps
meanTotalSteps<-as.numeric(format(meanTotalSteps,digits=5))
meanTotalSteps # round to integer 
```
Next we find the median total number of steps taken for each day.  
```{r median}
medianTotalSteps<- median(df_stepsPerDay$TotalSteps)
meanTotalSteps<-as.numeric(format(medianTotalSteps,digits=5))
medianTotalSteps # round to integer 
```
    
The mean total number of steps taken per day is 10766 (rounded to integer).  
The median total number of steps taken per day is 10765.  
  
## What is the average daily activity pattern?  

To plot the average number of steps per 5-minute interval:
```{r plot}
df_GroupByInterval<-group_by(df2,interval)
tail(df_GroupByInterval)
df_AveStepsInterval <- summarise(df_GroupByInterval,MeanSteps = mean(steps))
head(df_AveStepsInterval)
plot2<-qplot(interval,MeanSteps,data=df_AveStepsInterval,geom='point')
plot2 + geom_smooth(se=FALSE) + labs(x='interval',y='Average number of steps taken',
        title='Average Daily Activity Pattern')
```
  
To find the 5-minute interval where the average number of steps taken is at a maximum:
```{r IntervalOfMaxSteps,fig.height=4}
max <- max(df_AveStepsInterval$MeanSteps)
# interval at which max avg steps occur
maxInt<-df_AveStepsInterval[df_AveStepsInterval$MeanSteps==max,]$interval 
maxInt
# plot where max average step occurs
plot2 + geom_hline(yintercept = max) + geom_vline(xintercept = maxInt)
```  
  
  
Therefore the interval 835 contains the maximum average number of steps.  

## Imputing missing values
```{r MissingValues}
df_NA<-df[is.na(df$steps),]
nrow(df_NA)
```
There are 2304 missing values in the dataset.  

Next, fill in all missing values as mean of the corresponding 5-min interval.  

'impdata' is the new dataset with the missing values filled in.  
```{r StrategyMissing}
impData<-cbind(df,MeanSteps=df_AveStepsInterval$MeanSteps)
impData$steps[is.na(impData$steps)] <- impData[is.na(impData$steps),]$MeanSteps
head(impData)
```
Plot histogram of total number of steps taken each day, accounting for filled-in missing values. 
```{r NewHistogram,fig.height= 4}
Newdf_GroupByDate<-group_by(impData,date) # group by 'date' as factor
Newdf_stepsPerDay<-summarise(Newdf_GroupByDate,TotalSteps=sum(steps)) # create new data frame
plot3<-qplot(TotalSteps,data=Newdf_stepsPerDay,geom='histogram')
plot3+labs(x='Total Steps Taken per Day')
```
  
  
  Find new mean and median.  
```{r NewmeanAndMedian}
NewMeanTot<- mean(Newdf_stepsPerDay$TotalSteps)
NewMeanTot<-as.numeric(format(meanTotalSteps,digits=5))
NewMeanTot # round to integer 

NewMedTot<- median(Newdf_stepsPerDay$TotalSteps)
NewMedTot<-as.numeric(format(meanTotalSteps,digits=5))
NewMedTot 
```
Both the new mean and median steps taken are 10765.
The mean remains the same as before missing values are filled in, and the median
is increased by 1. Imputing estimates of total daily steps as the mean of corresponding 5-min interval 
slightly overestimates the median, but does not affect the mean in this scenario.   

## Are there differences in activity patterns between weekdays and weekends?
Create new factor variable comprised of the levels 'weekday' and 'weekend'. Incorporate into dataset with filled-in missing vlaues to differentiate date as weekday or weekend. 
```{r FactorVariable}
impData$date<-as.Date(impData$date)
WeekDays<-c('Monday','Tuesday','Wednesday','Thursday','Friday')
impData$DayType<-factor((weekdays(impData$date) %in% WeekDays), levels=c(FALSE,TRUE),labels=c('weekend','weekDay'))
df_WeekDay <- impData[impData$DayType == 'weekDay', ]
df_Weekend<- impData[impData$DayType == 'weekend', ]
```
We can now plot the average steps taken for both weekends and weekdays.
```{r PlotByDayType}
df_WeekDayInt<-group_by(df_WeekDay,interval)
head (df_WeekDayInt)
df_WeekdayAve <- summarise(df_WeekDayInt,MeanSteps = mean(steps))
head(df_WeekdayAve)
plot4 <- qplot(interval,MeanSteps,data=df_WeekdayAve,geom = 'line' )
plot4 + labs(x='interval',y='Average number of steps taken',
             title='Average Daily Activity Pattern for Weekdays')

df_WeekEndInt <- group_by(df_Weekend,interval)
tail(df_WeekEndInt)
df_WeekendAve <- summarise(df_WeekEndInt,MeanSteps = mean(steps))
head(df_WeekendAve)
plot5 <- qplot(interval,MeanSteps,data=df_WeekendAve,geom = 'line' )
plot5 + labs(x='interval',y='Average number of steps taken',
             title='Average Daily Activity Pattern for Weekends')
```
